---
title: 'Pre-Live-Session Coding Exercise 1'
author: "Dr. Jeffrey Yau"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
---

**DUE: 11:59pm Pacific Time on Monday, August 30.**

\newpage
# Question 1: Linear Probability Model

* **What are the advantages of the linear probability model?**
* **What are the drawbacks of the linear probability model?**

\newpage
# Question 2: Examining the dataset befor conducting EDA

Insert the function to *tidy up* the code when they are printed out
```{r}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## Practical Tips for Implementing Binary Logistic Regression 

When solving data science problems, always begin with the understanding of the underlying (business, policy, scientific, etc) question; our first step is typically **NOT** to jump right into the data.

In this example, suppose the question is *"Do females who have higher family income (excluding wife's income) have lower labor force participation rate?" If so, what is the magnitude of the effect?* Note that this was not objective in *Mroz (1987)'s* paper. For the sake of learning to use logistic regression in answering a specific question, we stick with this question in this example.

Understanding the sample data: Remember that this sample comes from *1976 Panel Data of Income Dynamics (PSID)*. PSID is one of the most popular datasets used by labor economists.

First, load the `car` library in order to use the Mroz dataset and understand the structure dataset. Once the `car` library is loaded, the `Mroz` library can be called simply by calling its name `Mroz`.

Typical questions you should always ask when examining a dataset include the following.

**Exercise: Write your own codes to answer the following questions:**

  - What is the number of variables and the number of observations in the Mroz dataset?
  - Are these variables sufficient for you to answer you questions?
  - If not, what other variables would you like to have? What impact (qualitatively) might not having these variables have on your models?
  - Are there any missing values (in each of the variables)? If so, how many missing values in each of the variables?
  - Are there any abnormal values in each of the variables in the raw data?


```{r}
# Import libraries
library(car)
library(dplyr)
library(Hmisc)

# Load the first 3 rows of the Mroz dataset
head(Mroz,3)

## YOUR CODE TO BE HERE
```
\newpage
# Question 3: Descriptive statistical analysis of the data

**Exercise: Conduct EDA on the Mroz Dataset**

```{r}
# YOUR CODE TO BE HERE
```

*As a best practice, we will need to incorporate insights generated from EDA on model specification. In what follows, we employ a very simple specification that uses all the variables "as-is", but the focus in this exericse is on how to interpret the coefficients.*

\newpage
# Question 4: Estimate a Binary Logistic Regression

**Exercises:**
* Estimate a Binary Logistic Regression
* Print the summary of the model results
* Intrepret the model results

* Dependent variable: lfp
** Explanatoryvariables: k5, k618, age, wc, hc, lwy, inc

```{r}
# YOUR CODE TO BE HERE
```








